Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-07-08 manuscript/manuscript.qmd
changed referencing style to CPT style
replaced observational with non-interventional

editor: visual
bibliography: references.bib
csl: journal-of-clinical-oncology.csl

@@

However, the comparability and transportability of results derived between observational studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of observational studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when observational studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, observational studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

@@

The focus of ENCORE is to efficiently explore in what contexts observational studies can or cannot yield similar results compared to oncology RCTs. Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years. After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).

@@

Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results. Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

Principled, well-designed and reproducible studies using fit-for-purpose RWD to generate RWE, such as through trial emulation, may complement evidence from RCTs. Through a systematic benchmarking approach, the ENCORE project will provide insights as to how methodological, design, and analytic decisions influence quantifiable bias and interpretability of observational studies in oncology.

>>>>

editor: visual
bibliography: references.bib
csl: clinical-pharmacology-and-therapeutics.csl

@@

However, the comparability and transportability of results derived between non-interventional studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of non-interventional studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when non-interventional studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, non-interventional studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

@@

The focus of ENCORE is to efficiently explore in what contexts non-interventional studies can or cannot yield similar results compared to oncology RCTs. Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years. After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).

@@

Causal analyses of non-interventional data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results. Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

Principled, well-designed and reproducible studies using fit-for-purpose RWD to generate RWE, such as through trial emulation, may complement evidence from RCTs. Through a systematic benchmarking approach, the ENCORE project will provide insights as to how methodological, design, and analytic decisions influence quantifiable bias and interpretability of non-interventional studies in oncology.
